\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2027}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath,amssymb}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}

\pdfinfo{/TemplateVersion (2027.1)}
\setcounter{secnumdepth}{0}

\title{When Do Failures Reveal Spurious Structure?\\A Study of Label-Free Latent-Risk Editing in Frozen Vision-Language Models}
\author{Anonymous Submission}
\affiliations{}

\begin{document}
\maketitle

\begin{abstract}
Errors of frozen vision-language models (VLMs) can concentrate in groups that are unobserved during adaptation. We ask whether task failures alone expose an editable latent-risk subspace without group or sensitive-attribute labels. \textsc{LatentRisk-VLM} converts task-labeled calibration margins into risk labels, removes class-level semantics, and estimates class-conditional risk directions by residual mean contrast or iterative linear probes. The subspace is defined operationally by failure predictiveness, weak semantic coupling, and suppressibility; it is not a discovered sensitive attribute. Experiments with frozen CLIP ViT-L/14 show a conditional rather than universal benefit. On Waterbirds, task-label-only rank-4 editing reaches 64.1 WGA in a three-seed rerun, while DFR-Frozen reaches 90.2 WGA with full attribute labels. On CelebA, rank-1 editing lowers WGA from 71.5 to 54.4 despite raising average accuracy from 78.2 to 91.7, while rank-4 worst-class selection reaches 72.8 WGA. Attribute-budget and concept-erasure experiments sharpen the interpretation: LatentRisk-VLM occupies the zero-attribute-label point, whereas explicit attribute labels remain highly beneficial when available. INLP raises Waterbirds WGA only to 43.1 and LEACE leaves WGA unchanged in our global feature-space erasure, and the measured risk--attribute subspace overlap is low. Additional low-cost extensions with frozen SigLIP and a MetaShift-style cat/dog contextual shift preserve the same cautionary conclusion: group-free task-accuracy selection can fail badly, while rank-4 worst-class selection can recover WGA only when the candidate pool and diagnostic evidence are favorable. Failure-derived directions are therefore useful weak supervision when failure and group shift align, but neither risk separability nor task accuracy guarantees group robustness.
\end{abstract}

\input{sections/1_introduction}
\input{sections/2_related_work}
\input{sections/3_latentrisk_vlm}
\input{sections/4_experimental_protocol}
\input{sections/5_measured_results_and_analysis}
\input{sections/8_low_cost_extension}
\input{sections/6_discussion_and_limitations}
\input{sections/7_conclusion}

\bibliography{latentrisk_vlm}

\end{document}
