\section{Experimental Protocol}
\paragraph{Backbone and reporting.}
The main experiments use 768-dimensional, normalized embeddings from the Transformers CLIP ViT-L/14 implementation. The low-cost extension additionally evaluates frozen SigLIP Base Patch16/224 \citep{zhai2023sigmoid}. In every case, both image and text encoders are frozen; the method operates only on cached shared-space embeddings. We report one deterministic run for each primary configuration. Results are measured adaptations to the stated frozen backbone, not numbers copied from the baseline papers.

\paragraph{Frozen-feature cost comparison.}
All methods in the controlled cost comparison share the same frozen CLIP backbone, the same cached image features, and the same calibration/evaluation split. Feature extraction time is reported separately from post-feature adaptation time. Under this shared frozen-backbone setting, we compare the additional supervision, optimization, and adaptation cost required after feature extraction. For controlled comparison, we adapt CnC and DFR to cached frozen CLIP features. These variants should not be interpreted as exact reproductions of the original end-to-end training procedures; they isolate the adaptation strategy under the same frozen representation. We therefore write CnC-Frozen and DFR-Frozen when referring to these feature-space variants, and ``-Frozen'' denotes feature-space adaptation using fixed CLIP embeddings.

\paragraph{Classification datasets.}
Waterbirds \citep{sagawa2020groupdro} uses the official validation split (1,199 images) for risk discovery and parameter selection and the test split (5,794 images) for evaluation. Fixed prompts are ``a photo of a landbird'' and ``a photo of a waterbird.'' Groups cross bird type with place. CelebA \citep{liu2015faceattributes} uses Blond Hair as the task and Male as the evaluation attribute. Its validation split (19,867 images) calibrates \textsc{LatentRisk-VLM}; its training split is available to fitted baselines; and its test split (19,962 images) is evaluated once. Prompts describe a person with or without blond hair. For the third classification setting, we build a deterministic MetaShift-style cat/dog contextual shift from GQA scene graphs \citep{hudson2019gqa} and Visual Genome images \citep{krishna2017visualgenome}, following the contextual-shift motivation of MetaShift \citep{liang2022metashift}. The group attribute is indoor versus outdoor context. The split contains 1,500 train, 300 validation/calibration, and 144 test images; validation and test are balanced over the four class-context cells.

\paragraph{Retrieval datasets.}
FairFace \citep{karkkainen2021fairface} supplies a 10,954-image validation gallery and the 25 gender-stereotype queries released with BEND-VLM. Its 86,744-image training split is used only as the gender-labeled reference bank for BEND-VLM. We report text-to-image retrieval on the Flickr30K \citep{young2014image} Karpathy test split \citep{karpathy2015deep}: 5,000 captions and 1,000 images. The 1,014-image validation split fits the LatentRisk retrieval editor. SANER is trained for five epochs on 124,026 person-related caption pairs from 27,576 disjoint Flickr30K training images.

\paragraph{Supervision accounting.}
We separate four classification settings according to where group attributes $a$ enter the pipeline. \emph{No example labels} includes frozen CLIP and prompt ensembling. \emph{Task labels only (group-free adaptation and selection)} includes ERM, JTT-style calibration, and the primary \textsc{LatentRisk-VLM} variants; these methods use $y$ but not $a$ for either fitting/adaptation or hyperparameter selection. \emph{Group-free adaptation, group-aware selection} includes CFR, CnC-Frozen, GIC, AFR, and our attribute-oracle selector: their reported fitting or editor construction does not use true group labels, but the operating point or checkpoint is selected with validation WGA computed from $(y,a)$. \emph{Group-aware adaptation and selection} includes SPD and BEND-VLM, which use attribute information during direction/reference construction and also retain attribute-aware validation. Group labels read only after prediction to compute test WGA do not affect this categorization. This accounting distinguishes learning an intervention without $a$ from selecting among interventions with $a$.

\paragraph{Baselines.}
Classification evaluates frozen CLIP, prompt ensembling, ERM, JTT \citep{liu2021jtt}, AFR \citep{qiu2023afr}, CnC-Frozen \citep{zhang2022correct}, BEND-VLM \citep{gerych2024bendvlm}, SPD \citep{zhao2026subspace}, CFR \citep{you2024calibrating}, and a GIC adaptation. The reviewer-risk extension additionally includes DFR-Frozen \citep{kirichenko2023dfr} when a feature-space classifier refit is available, and INLP/LEACE concept erasure on cached embeddings. Retrieval evaluates frozen CLIP, projection prompts \citep{chuang2023debiasing}, BEND-VLM, SANER \citep{hirota2025saner}, and \textsc{LatentRisk-VLM}. Where a reference method originally trains a different backbone, we retain its central objective and selection rule but operate on CLIP ViT-L/14 features or lightweight adapters. This creates a controlled frozen-backbone comparison, not an exact reproduction of each paper's original setting.

\paragraph{Hyperparameters.}
Classification uses $\tau=1$, $\eta=0.5$, full projection, and $\beta=0.5$. The measured CLIP rank-1 configuration uses mean contrast on Waterbirds and one logistic probe on CelebA; rank four uses iterative probes with $C=1$ and a 0.55 training-accuracy stopping threshold. Waterbirds searches 101 values of $\lambda\in[0,5]$ and 61 intercepts in $[-0.03,0.03]$; CelebA searches 51 values in $[0.05,0.30]$ over the same intercept grid. Rank-4 classification reports the full group-free selector sweep over this fixed candidate pool; the main group-free rank-4 row uses worst-class validation selection. The SigLIP extension keeps the same frozen-feature protocol, runs only frozen, rank-1, rank-4 worst-class, and rank-4 validation-WGA oracle settings, and reports WGA gains relative to frozen SigLIP. The MetaShift-style setting uses SigLIP, rank four, iterative probes, $\lambda\in[0.05,0.30]$, the same intercept grid, and group-free worst-class validation selection. Retrieval searches 61 values of $\lambda\in[0,0.3]$ and selects $0.1$. The Waterbirds ablation changes one component at a time under group-free, task-label-only macro-accuracy selection.

\paragraph{Diagnostic prediction protocol.}
For the new MetaShift-style shift, we generate the diagnostic prediction before held-out test editing. A calibration-only run writes risk scores, residuals, and risk subspaces, then computes split-half stability, risk concentration, and risk-probe AUC without test labels. Post-hoc validation alignment and minority enrichment use validation group labels only for diagnosis. A fixed YAML rule maps these diagnostics to \emph{likely success}, \emph{uncertain}, or \emph{likely failure}; the resulting \texttt{prediction.json} explicitly records that test performance was not used. Only after this file is written do we evaluate frozen SigLIP and final rank-4 editing on the test split.

\paragraph{Metrics.}
Classification reports average accuracy and WGA over the four $(y,a)$ groups. We also discuss their difference as the average--worst gap. Retrieval reports text-to-image Recall@10 and mean gender MaxSkew@100 over the 25 FairFace queries. MaxSkew compares retrieved gender proportions with a uniform target; lower is better. For the first two settings, group attributes are accessed only after prediction to compute audit metrics. The remaining settings use $a$ during selection or adaptation as stated explicitly in Table~\ref{tab:classification}.
