\section{Discussion and Limitations}
\paragraph{Failure--group alignment is the key assumption.}
The method never identifies a group as such. It identifies representation variation associated with low margins within a class. Worst-group performance improves only when that variation overlaps the factor defining the worst group and can be removed without erasing class evidence. The alignment diagnostic in Table~\ref{tab:alignment-diagnostics} makes this assumption measurable: Waterbirds has higher risk--spurious alignment and its top-risk set is minority-enriched, whereas CelebA has low gender-direction alignment and no minority enrichment. The SigLIP and MetaShift-style extensions reinforce the same caveat. High or moderate alignment can still be insufficient if candidate selection destroys class evidence or if split-half stability is low. CelebA therefore fails not because the risk direction is noisy, but because it is stable and points at other error structure. Accordingly, \textsc{LatentRisk-VLM} should be described as latent-risk editing, not unsupervised demographic debiasing.

\paragraph{Discovery and selection are distinct.}
Tables~\ref{tab:classification} and~\ref{tab:selector-sweep} separate group-label use during adaptation from group-label use during model selection. The CelebA oracle result is informative precisely because it is not a primary result: group labels are absent from subspace discovery in both the 72.8- and 82.2-WGA rows, yet validation-WGA selection still produces a 9.4-point WGA gain over the best group-free selector. Thus, eliminating group labels from representation discovery does not by itself yield a fully group-free pipeline. A deployment-relevant selector must also choose rank and scoring parameters without validation groups. Future work should target group-label-free robust selection, for example through stability across environments or tail-risk criteria, rather than assuming average accuracy will choose a group-robust edit.

\paragraph{Selection supervision and adaptation supervision should not be conflated.}
Methods such as AFR, CnC-Frozen, CFR, and GIC avoid true group labels in their fitting stage in the reported protocol but use validation groups to choose checkpoints or hyperparameters. SPD and BEND-VLM use attribute information earlier, during direction or reference construction. These are materially different supervision regimes, even though both can improve WGA. CnC-Frozen outperforms our CelebA oracle under group-aware validation, showing that reliable validation groups remain powerful when available. Latent-risk editing addresses an unnamed-nuisance setting and should not be presented as a replacement for methods that can legitimately use group supervision.

\paragraph{Complementary operating regime.}
\textsc{LatentRisk-VLM} targets a complementary operating regime: frozen vision-language models with task labels but without sensitive-attribute annotations. Our goal is not to replace fully supervised group-robust training when dense group annotations and retraining resources are available. Instead, we study whether model failures themselves can provide a low-cost signal for post-hoc risk suppression in frozen VLM representations. The Waterbirds budget experiment confirms that explicit attribute labels can be much stronger: DFR-Frozen reaches 90.2 WGA with the full validation attribute set, compared with 64.1 for LatentRisk-VLM-UL. This result should be reported directly. The relevant claim for LatentRisk-VLM is not dominance over attribute-supervised methods, but viability at the zero-attribute-label point and a diagnostic account of when failure supervision is informative.

\paragraph{Experimental limitations.}
The added SigLIP runs broaden the backbone evidence but do not establish architecture-level robustness: they cover one additional backbone, one deterministic seed for main results, and only the minimal frozen/rank-1/rank-4 matrix. The MetaShift-style cat/dog setting is constructed from GQA scene graphs and Visual Genome images with a deterministic keyword heuristic; it should not be treated as a full official MetaShift benchmark reproduction. The classification datasets remain binary, and the retrieval audit reduces gender to the binary labels supplied by FairFace and uses only 25 stereotype queries. The measured rank-1 estimator differs across the original CLIP classification datasets (mean contrast on Waterbirds and a logistic probe on CelebA), although the SigLIP estimator-control runs suggest that this choice is not the sole driver of the observed rank-1 failures. The alignment diagnostic itself uses group labels post hoc and is explanatory rather than deployable; the concentration and stability diagnostics are group-free but are still measured on a small number of settings. Several baselines are adaptations to frozen CLIP features rather than exact reproductions in their original architectures and data regimes. The same validation split is used for risk discovery and parameter selection, which can favor calibration-specific structure. Finally, a predictive direction is correlational: these experiments do not establish that it causally generates errors or that editing it is safe under unmeasured shifts.

\paragraph{Ethical use.}
An attribute-free method can still worsen a protected group, as the CelebA result demonstrates. Aggregate accuracy must not be used to certify fairness, and post-hoc subgroup audits remain necessary whenever they are appropriate and lawful. The method should not be deployed to remove demographic information indiscriminately or to infer sensitive attributes. Its appropriate role is as an auditable representation intervention whose group consequences are measured explicitly.
